---
title: "Championship_Predictor"
author: "Ryan Abbott"
date: "`r Sys.Date()`"
output:
  word_document: default
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This project will attempt to predict the winner of the NBA Finals based on several team-based statistics
```{r}
Sys.setenv("VROOM_CONNECTION_SIZE" = 262144)
library("nbastatR")
library("dplyr")
library("tidyverse")
```

```{r}
#game_logs(seasons = 2023)
#get_day_nba_league_standings("02/1/2023")
#get_team_season_stat_splits(year.season_start  = 2015)
teamdata <- tibble(read.csv("training_data_team_raw.csv"))
#teamdata <- na.omit(teamdata$ptsPerGameDiff)
teamdata
```

## Exploratory Data Analysis: Finding variables

First, I'm thinking that point differentials between a team and their opponents are important to overall success. Lets visualize that.

```{r}
df <- teamdata %>% select(ptsPerGameDiff, slugSeason, nameTeam, recordOverall, pctWinTeam, nameConference, wins) %>% arrange(desc(ptsPerGameDiff))

plot <- ggplot(df, aes(x=ptsPerGameDiff,y=pctWinTeam,col=nameConference))

plot <- plot + geom_point()

plot
```

The correlation between these two variables is astoundingly close. The model can be better optimized though.



```{r}
ppg <- teamdata %>% select(ptsPerGameTeam, ptsPerGameOpponent, nameTeam, slugSeason, pctWinTeam, wins, nameConference, recordOpponentOver500WinPct, ptsPerGameDiff, recordAheadAtThirdWinPct, recordAwayWinPct, recordAheadAtHalfWinPct, recordHomeWinPct) %>% arrange(desc(ptsPerGameTeam))

plot <- ggplot(ppg, aes(x=ptsPerGameTeam, y=wins, col=nameConference)) + geom_point()
plot2 <- ggplot(ppg, aes(x=ptsPerGameOpponent, y=wins, col=nameConference)) + geom_point()
plot3 <- ggplot(ppg, aes(x=recordOpponentOver500WinPct, y=wins, col=nameConference)) + geom_point()
plot4 <- ggplot(ppg, aes(x=recordAheadAtHalfWinPct, y=wins, col=nameConference)) + geom_point()
plot5 <- ggplot(ppg, aes(x=recordAwayWinPct, y=wins, col=nameConference)) + geom_point()
plot6 <- ggplot(ppg, aes(x=recordHomeWinPct, y=wins, col=nameConference)) + geom_point()


plot
plot2
plot3
plot4
plot5
plot6
```

Most of these results do not surprise me. Points scored per game is a very relative statistic, as a team with slower pacing may have dozens of possessions less than a team with faster pacing. The same applies for points allowed per game.

A team's record against other teams with a winning record strongly correlates to a better win percentage as well.

One statistic that I am a bit surprised correlates so strongly with total wins is a team's record when they lead the game at halftime. I expected a relatively strong positive correlation, but with half of a game still remaining to play, it is easy to envision that the opposing team can mount a comeback.



## Model Testing
```{r}
champion.lm = lm(wins~ptsPerGameDiff+recordOpponentOver500WinPct+recordAheadAtThirdWinPct+recordAheadAtHalfWinPct+recordAwayWinPct, ppg)
```


```{r}
summary(champion.lm)
```

The following code chunk is just to check the normality of the distribution. I may perform some probability analysis at a later time, but for now, this is all I have done.

Evidently, the distribution is not normal when looking at the lesser teams.
```{r}
library(s20x)
normcheck(champion.lm,shapiro.wilk = TRUE)
cooks20x(champion.lm)
```

Here is the scatter plot for the final linear model. It is remarkably accurate, with some teams performing slightly worse than predicted but virtually zero performing better than expected.
```{r}
pred <- function(x1, x2, x3, x4, x5) {
  champion.lm$coefficients[1] + champion.lm$coefficients[2]*x1 + champion.lm$coefficients[3]*x2 + champion.lm$coefficients[4]*x3 + champion.lm$coefficients[5]*x4 + champion.lm$coefficients[6]*x5
}

x=pred(ppg$ptsPerGameDiff,ppg$recordOpponentOver500WinPct,ppg$recordAheadAtThirdWinPct,ppg$recordAheadAtHalfWinPct,ppg$recordAwayWinPct)

plot(wins~x, ppg)
```

## Model Testing: 2022 season

```{r}
team_data_test <-  standings(seasons = 2022)
```
```{r}
team_data_test
ppg_test <- team_data_test %>% select(slugSeason, nameTeam, wins) #%>% arrange(desc(ptsPerGameDiff))

test <-  team_data_test %>% select(recordOpponentOver500WinPct,ptsPerGameDiff,recordAheadAtThirdWinPct,recordAheadAtHalfWinPct,recordAwayWinPct)

ppg_test <- ppg_test %>% mutate(projected_wins = c(pred(test$ptsPerGameDiff, test$recordOpponentOver500WinPct, test$recordAheadAtThirdWinPct, test$recordAheadAtHalfWinPct, test$recordAwayWinPct)))

arrange(ppg_test, desc(projected_wins))
```

## NBA Teams' Projected Wins, 2023
```{r}
team_data_2023 <- standings(seasons = 2023)
```

```{r}
ppg_test <- team_data_2023 %>% select(slugSeason, nameTeam, wins)


ppg_test <- ppg_test %>% mutate(projected_wins = c(pred(team_data_2023$ptsPerGameDiff, team_data_2023$recordOpponentOver500WinPct, team_data_2023$recordAheadAtThirdWinPct, team_data_2023$recordAheadAtHalfWinPct, team_data_2023$recordAwayWinPct)))

arrange(ppg_test, desc(projected_wins))
```

